Reinforcement Learning Methods for Operations Research Applications -- Reloaded

### 2020-12-18 paper/paper.tex
update

Therefore, most applications that incorporate and directly reflect costs or profit in the reward
function use average reward reinforcement learning. \citet{aydin2000dynamic} use it in the setting
of scheduling, while in a series of papers Mahadevan et al. investigated several problem domains
starting with simple MDPs
\citep{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults,Mahadevan96_OptimalityCriteriaInReinforcementLearning}.
After these foundational works for average reward reinforcement learning they introduced a
continuous time average reward reinforcement learning algorithm named SMART
\citep{Mahadevan97_SelfimprovingFactorySimulationUsingContinuoustimeAveragerewardReinforcementLearning}.
Applications of SMART reach from the optimisation of queuing systems
\citep{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies,Mahadevan96_SensitiveDiscountOptimalityUnifyingDiscountedAndAverageRewardReinforcementLearning},
maintenance of an inventory system
\citep{Das99_SolvingSemiMarkovDecisionProblemsUsingAverageRewardReinforcementLearning} to optimising
transfer line in terms of maximizing demand, while keeping inventory levels of unfinished product as
low as possible \citep{Mahadevan98_OptimizingProductionManufacturingUsingReinforcementLearning}.
However, in practise usually decision have to be made on a daily basis
\citep{enns2004work,GeldersvanW1982}. Therefore, we refrain from this adaption and concentrate on
standard MDPs only. Furthermore, often continuous-time semi-MDP problems can be converted through
uniformisation into equivalent discrete time instances
\citep[see][]{Puterman94,bertsekas1995dynamic}.

Like discounted reinforcement learning also average reward reinforcement learning is based on an
oracle function, in our case the accumulated costs of a period, to assess the decisions taken by the
agent. By repeatedly choosing different actions the agent examines the problem space and rates
possible actions for any observed state. The advantage of average reward reinforcement learning over
the widely applied discounted reinforcement learning framework is that the underlying optimisation
technique is able to find better policies. This yields from the fact that in standard discounted
reinforcement learning method the states are assessed independently and by a single value. In
contrast to that average reward reinforcement learning splits up the evaluation of the average
reward per step, a bias value that specifies the amount of collected rewards to reach the optimal
path when starting in a suboptimal state and an error term which defines the number of steps to
reach the optimal path
\citep{Howard64,Puterman94,Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults}.
Thus, while in average reward reinforcement learning these terms are learned separately in the
discounted framework one value consisting of the addition of these values is estimated, where
however the average reward is scaled by \(1/(1-\gamma)\) with the discount factor \(\gamma\) usually
being set very close to \(1\), e.g. \(0.99\). In commonly applied unichain MDPs the average reward
per step is equal for all states. Thus, independently assessing it is computationally unwisely.
Therefore, we adapt an algorithm that uses a scalar value for the estimation of the average reward.
Furthermore, as the average reward is scaled in the discounted state values it dominates the other
two terms. This, combined with the iterative evaluation and the independently assessing of the
average reward for each state lets standard discounted reinforcement learning struggle to find good
policies.
% Thus, this incautious combining of different kinds of state values as done in discounted
% reinforcement learning leads to the problems that (i) the average reward, which is equal for all
% states of usually investigated unichain MDPs, is dominating and thus diluting the bias values, and
% (ii) the state values are deteriorated by the error term that is only imposed due to the
% discounting technique.

Thus, as opposed to the commonly applied discounted reinforcement learning algorithm we use an
average reward adjusted reinforcement learning algorithm to adaptively release orders based on the
assessed state values of the production system. In contrast to the aforementioned works on average
reward reinforcement learning our algorithm incorporates the optimisation of not only the average
reward over time, but also the bias values, which is an important adaption in highly stochastic
systems. Only the work by
\cite{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies}
integrates this second-level refinement optimisation. However, their algorithm requires the
selection of a reference state to prevent an infinite increase of state values. This results from
the lack of feedback in the iterative process of optimising the different refinement optimisation
levels. Furthermore, they asses the average reward independently.

Therefore, we propose a novel average reward reinforcement learning algorithm, which is the first
that solves the occurring cyclic constraint problem in the setting of average reward reinforcement
learning, to assess orders for their release to the shop floor. These cyclic constraint problem
emerges as the underlying constraint structure is based on the Laurent series expansion of the state
values as shown by \cite{MillerVeinott1969}, and thus easily imposes an infinite number of
interconnected constraints when handled unwisely.
%
The agent assesses the expected costs for the possible releases imposed by adapting the lead time.
According to the agent's estimates it sets a planned lead time and with that releases orders into
the production system.
% upon releasing an order for the current and future periods
% and, based on its estimates decides whether to release the order or not.
%
% \MS[t]{TODO} We provide ample evidence of the viability of the approach and compare it to well-known
% order release mechanisms. Furthermore, we show that the advantage of average reward reinforcement
% learning for the problem structures encountered in the context of production and logistics by
% comparing the results to the discounted reinforcement learning framework. The preliminary results
% show that our approach performs better than the other methods. With regard to practical implications
% we are confident that decision support tools based on average reward reinforcement learning increase
% the decision quality of human planners.

@@

Thus, the action space is defined as \(F = \times_{s=1}^{\size{\States}} A_{s}\), where \(A_{s}\) is a
finite set of possible actions. A \emph{policy} is a sequence \(\pol = (f_{1},f_{2},\ldots)\) of
elements \(f_{t} \in F\). Using the policy \(\pol\) means that if the system is in state \(s\) at
time \(t\) the action \(f_{t}(s)\), i.e.% that is
the \(s\)-th component of \(f_{t}\), is chosen. A stationary policy \(\pol = (f,f,\ldots)\) does not
depend on time. In the sequel we are concerned with stationary policies only. Thus the goal in
reinforcement learning (RL) is to find the best stationary policy \(\pol\) for a given problem,
where the phrase ``best'' will be elaborated in detail below.

@@

The aim in the discounted framework is to find an optimal policy \(\polopt_{\gamma}\), which when
followed, maximises the state value for all states \(s\) as compared to any other policy
\(\pol_{\gamma}\): \(V_{\gamma}^{\polopt_{\gamma}} - V_{\gamma}^{\pol_{\gamma}} \geqslant 0\). This
criteria is usually referred to as \(\gamma\)-optimality as the discount factor \(\gamma\) is fixed.
However, this also means that the actual value set for \(\gamma\) defines the policy that is
optimised for. For instance, as can be seen in Figure~\ref{fig:printer} setting \(\gamma\) value to
\(<0.8027\) the printer-loop is preferred over the mail-loop by the agent, although the mail-loop
accumulates more reward over time, i.e. selecting the mail-loop is a sub-optimal choice.

\begin{figure}[t!]
  \centering
  \begin{tikzpicture}[thin, scale=0.75]
    % Nodes
    \foreach \x in {1,...,5}{
      \draw ({cos(\x*72-72)*2.5},{sin(\x*72-72)*2.5}) node(\x) [circle,draw,minimum size=25] {\footnotesize \(\x\)};
    };
    \foreach[evaluate={
      \y=int(\x+5);
    }]  \x in {2,...,3}{
      \draw ({-cos(\x*72-72)*2.5+5},{sin(\x*72-72)*2.5}) node(\y) [circle,draw,minimum size=25]
      {\footnotesize \(\x'\)};
    };
    \foreach [evaluate={
      \y=int(\x+5);
      \z=int(\x+5);
    }] \x in {4,...,5}{
      \draw ({-cos(\x*72-72)*2.5+5},{sin(\x*72-72)*2.5}) node(\z) [circle,draw,minimum size=25]
      {\footnotesize \(\y'\)};
    };

    % Edges
    \path[thin, ->, bend right, >=stealth] (1) edge[below left] node {\footnotesize\(\)} (2);
    \path[thin, ->, bend right, >=stealth] (2) edge[above]      node {\footnotesize\(\)} (3);
    \path[thin, ->, bend right, >=stealth] (3) edge[left ]      node {\footnotesize\(\)} (4);
    \path[thin, ->, bend right, >=stealth] (4) edge[above]      node {\footnotesize\(\)} (5);
    \path[thin, ->, bend right, >=stealth] (5) edge[above left] node {\footnotesize\(5\)} (1);

    \path[thin, ->, bend left, >=stealth]         (1)  edge[below right] node {\footnotesize\(\)} (7);
    \path[thin, ->, bend left, >=stealth]         (7)  edge[above right] node {\footnotesize\(\)} (8);
    \path[thin, ->, dashed, bend left, >=stealth] (8)  edge[right]       node {\(\ldots\)} (9) ;
    \path[thin, ->, bend left, >=stealth]         (9)  edge[above]       node {\footnotesize\(\)} (10);
    \path[thin, ->, bend left, >=stealth]         (10) edge[above right] node {\footnotesize\(20\)} (1);

    \draw (0,0) node[] { Printer };
    \draw (5,0) node[] { Mail };

  \end{tikzpicture}
  \caption{\label{fig:printer} A MDP with two different deterministic policies. The only action choice is in
    state \(1\), in which the agent can choose between doing the printer-loop or the mail-loop.
    Observe that the average reward received per step equals \(1\) for the printer-loop and \(2\)
    for the mail-loop. Thus, the (Blackwell-)optimal policy is to choose the mail-loop. However,
    if \(\gamma < 3^{-\frac{1}{5}} \approx 0.8027\) an agent using discounted reinforcement
    learning chooses the printer loop.
    %
    Adapted from \cite{Mahadevan96_OptimalityCriteriaInReinforcementLearning}.}
\end{figure}

One idea behind the \(\gamma\)-parameter of the discounted framework is to be able to balance
short-term rewards (low \(\gamma\)-values) and long-term rewards (high \(\gamma\)-values). However,
what seems to be an advantage rather becomes a disadvantage for most applications. The issue is that
the average reward value is non-linearly increased when \(\gamma\) approaches one. However, in
almost all cases the aim is to perform well over time which in terms of reward means to first
maximise for a the policy with highest average reward before more selectively choosing actions. That
is, if several processes exist, we are searching for the ones with highest average reward before
considering other criteria.
%
Therefore, in almost all RL studies the discount factor is set to a value very close to \(1\), for
instance to \(0.99\)~\cite[e.g.]{mnih2015human,mnih2016asynchronous,Lillicrap15}.
%
By doing so the above mentioned non-linear relationship leads to the fact that the state value
consists almost solely of the up-scaled average reward term, whereas the bias values can be
neglected. This leads to diluted state values and thus difficulties in distinguishing actions which
impose policies that possess the same average reward.
%
In contrast to that, average reward RL separately assesses these values and according to these
selectively chooses the best action, comparing one after the other.
%
To illustrate the idea reconsider Figure~\ref{fig:printer}. First the agent picks the policy
according to the highest average reward. Thus, all actions are assessed by an average reward of
\(2\) which is inferred from the Mail-loop. This however, makes choosing the Printer-loop
unattractive; i.e. the policy converges to the optimal policy of choosing the Mail loop.

@@

In the common case of unichain MDPs, in which only a single set of recurrent states exists, the
average reward \(\avgrew^{\pol}(s)\) is equal for all states \(s\)
\citep{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults,Puterman94}.
For simplicity we concentrate on unichain MDPs in this work and thus may simply refer to it as
\(\avgrew^{\pol}\). A policy that maximises the average reward in every state is called
gain-optimal. Gain optimality is the least selective criteria an average reward reinforcement
learning aims for.

@@

where again \(R_{t}^{\pol}(s)\) is the reward received at time \(t\), starting in state \(s\) and
following policy \(\pol\). Note that the bias values are bounded due to the subtraction of the
average reward. Thus the bias value can be seen as the rewards that additionally sum up in case the
process starts in state \(s\). A policy that is gain-optimal is also bias-optimal if it maximises
the bias values in every state and compared to every other policy.

Especially for highly probabilistic systems bias-optimality is important. To clarify this consider
Figure~\ref{fig:three-states} which again consists of two possible deterministic policies with the
only choice in state \(1\). Both policies have the same average reward of \(1\). However, only
selecting the A-loop is bias-optimal, as the actions with non-zero rewards are selected earlier.
E.g.~consider starting in state 1. Under the policy \(\pol_{A}\) which takes the A-loop the reward
sequence is \((2,0,2,0,\ldots)\), while for the other policy \(\pol_{B}\) it is
\((0,2,0,2,\ldots)\).

@@

  \end{tikzpicture}
  \caption{\label{fig:three-states} A task with two different MDPs, A (going left in 1) and B
    (going right). As in the previous example the only action choice is in state \(1\). Observe
    that the average reward received per step equals \(1\) for both policies. However, only taking
    the A loop is bias-optimal, as its policy \(\pol_{A}\) leads to bias values
    \(V^{\pol_{A}}(1)=0.5\), \(V^{\pol_{A}}(0)=-0.5\) and \(V^{\pol_{A}}(2)=1.5\), while policy
    \(\pol_{B}\) which selects the B loop generates bias values \(V^{\pol_{B}}(1)=-0.5\),
    \(V^{\pol_{B}}(0)=-1.5\) and \(V^{\pol_{B}}(2)=0.5\). Adapted from
    \cite{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults}.}
\end{figure}

\subsection{More Refined Optimality Criteria}
\label{subsec:More_Refined_Optimality_Criteria}

The aforementioned and informally introduced notions of gain-optimality and bias-optimality can be
generalised to \(n\)-discount-optimality and will now be defined formally.

\begin{definition}
  Due to \cite{Veinott69} for MDPs a policy \(\polopt\) is \emph{\(n\)-discount-optimal} for
  \(n=-1,0,1,\ldots\) for all states \(s \in \States\) with discount factor
  \(\gamma\) % and \(V_{\gamma}^{\pol}(s)\) being the value function as defined above
  %
  if and only if
  \begin{align*}
    \lim_{\gamma \to 1}(1-\gamma)^{-n}\ (V_{\gamma}^{\polopt}(s) - V_{\gamma}^{\pol}(s)) \geqslant 0 \tpkt
  \end{align*}
\end{definition}

For the case of \(n=-1\) this lead to gain-optimality
\(\avgrew^{\polopt}(s) - \avgrew^{\pol}(s) \geqslant 0\) for all policies \(\pol\) and states \(s\)
\citep{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies}. In
the case of \(n=0\) it describes bias-optimality with \(V^{\polopt}(s) - V^{\pol}(s) \geqslant 0\)
forall policies \(\pol\) and states \(s\)
\citep{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies}.
%
Only if a policy is \(n\)-discount-optimal for all \(n < m\) it can be \(m\)-discount
optimal \citep{Puterman94,Veinott69}.
%
If a policy is \(\infty\)-discount-optimal then it is said to be
Blackwell-optimal \citep{Blackwell62}.
% weakening the argument!!!
I.e.~for Blackwell-optimal policies \(\polopt\) there exists a discount factor \(\gammaopt < 1\)
such that \(V_{\gamma}^{\polopt}(s) \geqslant V_{\gamma}^{\pol}(s)\) for all
\(\gamma \geqslant \gammaopt\) and under all policies \(\pol\)
\citep{Mahadevan96_SensitiveDiscountOptimalityUnifyingDiscountedAndAverageRewardReinforcementLearning,Blackwell62}.
Informally that means there exists a discount factor \(<1\) which finds Blackwell-optimal policies.
However, (i) in more complex, i.e. real world, MDPs this value can be arbitrary close to \(1\) and
(ii) the difference in state values may be very small, which (iii) due to the need of state value
function approximation likely causes errors when choosing among different actions in the discounted
framework.
% Real world applications underlie much more complex MDPs as the ones of Figures~\ref{fig:printer}
% and~\ref{fig:three-states}. Thus \(\gamma\)-values even very close to \(1\) can lead to sub-optimal
% policies. To investigate these differences we will now present a more elective optimality criteria.
%
%

@@

where \cite{Puterman94} shows that \(\lim_{\gamma \to 1} e_{\gamma}^{\pol}(s) = 0\). Note how the
first term depending on the average reward \(\avgrew^{\pol}(s)\) converges to infinity as \(\gamma\)
increases.

However, an important insight is the connection between \(n\)-discount-optimality and the Laurent
series expansion of \(V_{\gamma}^{\pol}(s)\). Each addend corresponds to one step in
\(n\)-discount-optimality. That is for \((-1)\)-discount-optimality the average reward must be
maximised.
%
Then for \(0\)-discount-optimality the agent has to choose the highest bias values.
%
Finally \(n\)-discount-optimality for \(n \geqslant 1\) requires to maximise the term
\(e_{\gamma}^{\pol}(s)\) which only exists as \(\gamma\) is strictly less than \(1\). Thus this term
incorporates the number of steps its expected rewards on the path which are required for reaching
the highest rewards. That is, it optimises the expected reward according to the occurrence on the
paths, where shorter paths and those which collect the rewards sooner are preferred.

>>>>

Therefore, most applications that incorporate and directly reflect costs or profit in the reward
function use average reward reinforcement learning. \citet{aydin2000dynamic} use it in the setting
of scheduling, while in a series of papers Mahadevan et al.\@ investigated several problem domains
starting with simple MDPs
\citep{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults,Mahadevan96_OptimalityCriteriaInReinforcementLearning}.
After these foundational works for average reward reinforcement learning they introduced a
continuous time average reward reinforcement learning algorithm named SMART
\citep{Mahadevan97_SelfimprovingFactorySimulationUsingContinuoustimeAveragerewardReinforcementLearning}.
Applications of SMART reach from the optimisation of queuing systems
\citep{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies,Mahadevan96_SensitiveDiscountOptimalityUnifyingDiscountedAndAverageRewardReinforcementLearning},
maintenance of an inventory system
\citep{Das99_SolvingSemiMarkovDecisionProblemsUsingAverageRewardReinforcementLearning} to optimising
transfer line in terms of maximizing demand, while keeping inventory levels of unfinished product as
low as possible \citep{Mahadevan98_OptimizingProductionManufacturingUsingReinforcementLearning}.
However, in practise usually decision have to be made on a daily basis
\citep{enns2004work,GeldersvanW1982}. Therefore, we refrain from this adaption and concentrate on
standard MDPs only. Furthermore, often continuous-time semi-MDP problems can be converted through
uniformisation into equivalent discrete time instances
\citep[see][]{Puterman94,bertsekas1995dynamic}.

Like discounted reinforcement learning also average reward reinforcement learning is based on an
oracle function, in our case the accumulated costs of a period, to assess the decisions taken by the
agent. By repeatedly choosing different actions the agent examines the problem space and rates
possible actions for any observed state. The advantage of average reward reinforcement learning over
the widely applied discounted reinforcement learning framework is that the underlying optimisation
technique is able to find better policies. This yields from the fact that in standard discounted
reinforcement learning method the states are assessed independently and by a single value. To be
more precise, average reward reinforcement learning splits up the evaluation of the average reward
per step, a bias value that specifies the amount of collected rewards to reach the optimal path when
starting in a suboptimal state and an error term which defines the number of steps to reach the
optimal path \citep{Howard64,Puterman94,
  Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults}. % Thus,
% while in average reward reinforcement learning these terms are learned separately
In contrast to that in the standard discounted framework one single scalar value consisting of the
addition of these subterms is estimated, where however the average reward is scaled by
\(1/(1-\gamma)\). The discount factor \(\gamma\) is usually set very close to \(1\), e.g. \(0.99\),
which leads to the fact that this subterm dominates the other two.
%
Further, in commonly applied unichain (and thus ergodic) MDPs the average reward per step is equal
for all states. Thus, independently assessing it is not only computationally unwisely, but also
problematic when iteratively annealed as done in reinforcement learning. Therefore, we adapt an
algorithm that uses a scalar value for the estimation of the average reward over all states, and
estimates for every state-action pair that incorporate the bias and error term. The later values are
adjusted by the average reward.
% This, combined with the iterative evaluation and the independently assessing of the average reward
% for each state lets standard discounted reinforcement learning struggle to find good policies.
% Thus, this incautious combining of different kinds of state values as done in discounted
% reinforcement learning leads to the problems that (i) the average reward, which is equal for all
% states of usually investigated unichain MDPs, is dominating and thus diluting the bias values, and
% (ii) the state values are deteriorated by the error term that is only imposed due to the
% discounting technique.

Thus, as opposed to the commonly applied discounted reinforcement learning algorithm we use an
average reward adjusted reinforcement learning algorithm to adaptively release orders based on the
assessed state values of the production system. In contrast to the aforementioned works on average
reward reinforcement learning our algorithm incorporates the optimisation of not only the average
reward over time, but also the bias values, which is an important adaption in highly stochastic
systems. Only the work by
\citet{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies}
integrates this second-level refinement optimisation. However, their algorithm requires the
selection of a reference state to prevent an infinite increase of state values. This results from
the lack of feedback in the iterative process of optimising the different refinement optimisation
levels. Furthermore, they asses the average reward independently for each state.

In summary, we propose a novel average reward adjusted reinforcement learning algorithm to assess
orders for their release to the shop floor. This is the first average reward adjusted reward
learning algorithm that operates using a artificial neural network as function approximation. To
ensure scalability we adapt a multi-agent approach, where each agent optimises the lead time
management of a single product type. The agents do so by assessesing the expected costs for the
possible releases imposed by adapting the lead time. According to the learned estimates each agent
sets a planned lead time for the corresponding product type and with that releases orders into the
production system. The highly optimised value-iteration algorithm uses improved n-step reinforcement
learning with small action-based replay memories and worker agents to infer order release policies
which outerperform \MS[t]{todo}.

%
% \MS[t]{TODO} We provide ample evidence of the viability of the approach and compare it to well-known
% order release mechanisms. Furthermore, we show that the advantage of average reward reinforcement
% learning for the problem structures encountered in the context of production and logistics by
% comparing the results to the discounted reinforcement learning framework. The preliminary results
% show that our approach performs better than the other methods. With regard to practical implications
% we are confident that decision support tools based on average reward reinforcement learning increase
% the decision quality of human planners.

@@

Thus, the action space is defined as \(F = \times_{s=1}^{\size{\States}} A_{s}\), where \(A_{s}\) is
a finite set of possible actions. A \emph{policy} is a sequence \(\pol = (f_{1},f_{2},\ldots)\) of
elements \(f_{t} \in F\). Using the policy \(\pol\) means that if the system is in state \(s\) at
time \(t\) the action \(f_{t}(s)\), i.e.% that is
the \(s\)-th component of \(f_{t}\), is chosen. A stationary policy \(\pol = (f,f,\ldots)\) does not
depend on time. In the sequel we are concerned with stationary policies only. The goal in
reinforcement learning (RL) is to find the best stationary policy \(\pol\) for a given problem,
where the phrase ``best'' will be elaborated in detail below.

@@

The aim in the discounted framework is to find an optimal policy \(\polopt_{\gamma}\), which when
followed, maximises the state value for all states \(s\) as compared to any other policy
\(\pol_{\gamma}\): \(V_{\gamma}^{\polopt_{\gamma}} - V_{\gamma}^{\pol_{\gamma}} \geqslant 0\). This
criteria is usually referred to as \(\gamma\)-optimality as the discount factor \(\gamma\) is fixed.
However, this also means that the actual value set for \(\gamma\) determines the best achievable
policy. For instance, as can be seen in Figure~\ref{fig:printer} setting \(\gamma\) value to
\(<0.8027\) the printer-loop is preferred over the mail-loop by the agent, although the mail-loop
accumulates more reward over time, i.e.\@ selecting the mail-loop is a sub-optimal choice. Observe
that the average reward received per step equals \(1\) for the printer-loop and \(2\) for the
mail-loop. Thus, the (Blackwell-)optimal policy is to choose the mail-loop. However, if
\(\gamma < 3^{-\frac{1}{5}} \approx 0.8027\) an agent using discounted reinforcement learning
chooses the printer loop.

\begin{figure}[t!]
  \centering
  \input{figures/printer}
  \caption{\label{fig:printer} A MDP with a single action choice in
    state \(1\), i.e.\@ two different deterministic policies
    % , in which the agent can choose between doing the printer-loop or the mail-loop  . The
    % Observe that the average reward received per step equals \(1\) for the printer-loop and \(2\)
    % for the mail-loop. Thus, the (Blackwell-)optimal policy is to choose the mail-loop. However,
    % if \(\gamma < 3^{-\frac{1}{5}} \approx 0.8027\) an agent using discounted reinforcement
    % learning chooses the printer loop.
    % %
    \citep[][adapted]{Mahadevan96_OptimalityCriteriaInReinforcementLearning}.
  }
\end{figure}

One idea behind the \(\gamma\)-parameter of the discounted framework is to be able to balance
short-term rewards (low \(\gamma\)-values) and long-term rewards (high \(\gamma\)-values). However,
what seems to be an advantage rather becomes a disadvantage for most applications. The issue is that
the average reward value is non-linearly increased when \(\gamma\) approaches one. However, in
almost all cases the aim is to perform well over time which in terms of reward means to first
maximise for a policy with highest average reward before more selectively choosing actions. That is,
we are searching for the policies with highest average reward before considering other criteria.
%
Therefore, in almost all RL studies the discount factor is set to a value very close to \(1\), for
instance to \(0.99\)~\citep[e.g.][]{mnih2015human,mnih2016asynchronous,Lillicrap15}.
%
By doing so the above mentioned non-linear relationship leads to the fact that the state value
consists almost solely of the up-scaled average reward term, whereas the bias values can be
neglected. This leads to diluted state values and thus difficulties in distinguishing actions which
impose policies that possess the same average reward.
%
In contrast to that, average reward RL separately assesses these values and according to these
selectively chooses the best action, comparing one after the other.
%
To illustrate the idea reconsider Figure~\ref{fig:printer}. First the agent picks the policy
according to the highest average reward. Thus, all actions are assessed by an average reward of
\(2\) which is inferred from the Mail-loop. This however, makes choosing the Printer-loop
unattractive; i.e. the policy converges to the optimal policy of choosing the Mail loop.

@@

In the common case of unichain MDPs, in which only a single set of recurrent states exists, the
average reward \(\avgrew^{\pol}(s)\) is equal for all states \(s\)
\citep{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults,Puterman94}.
In the sequel we focus on unichain MDPs in this work and thus may simply refer to it as
\(\avgrew^{\pol}\). A policy that maximises the average reward in every state is called
gain-optimal. Gain-optimality is the least selective criteria an average reward reinforcement
learning aims for.

@@

where again \(R_{t}^{\pol}(s)\) is the reward received at time \(t\), starting in state \(s\) and
following policy \(\pol\). Note that the bias values are bounded due to the subtraction of the
average reward. Thus the bias value can be seen as the rewards that additionally sum up in case the
process starts in state \(s\). A policy that is gain-optimal is also bias-optimal, if it maximises
the bias values in every state and compared to every other policy.

Especially for highly probabilistic systems bias-optimality is important. To clarify this consider
Figure~\ref{fig:three-states} which again consists of two possible deterministic policies with the
only choice in state \(1\). Both policies have the same average reward of \(1\). However, only
taking the A loop is bias-optimal, as its policy \(\pol_{A}\) leads to bias values
\(V^{\pol_{A}}(1)=0.5\), \(V^{\pol_{A}}(0)=-0.5\) and \(V^{\pol_{A}}(2)=1.5\), while policy
\(\pol_{B}\) which selects the B loop generates bias values \(V^{\pol_{B}}(1)=-0.5\),
\(V^{\pol_{B}}(0)=-1.5\) and \(V^{\pol_{B}}(2)=0.5\).
%
This yields from the fact that the actions with non-zero rewards are selected earlier in policy
\(\pol_{A}\). Consider starting in state 1. Under the policy \(\pol_{A}\) the reward sequence is
\((2,0,2,0,\ldots)\), while for the other policy \(\pol_{B}\) it is \((0,2,0,2,\ldots)\).

@@

  \end{tikzpicture}
  \caption{\label{fig:three-states} A MDP with two gain-optimal deterministic policies, \(\pol_{A}\)
    going left in 1 is also bias-optimal, while \(\pol_{B}\) is not \citep[Adapted
    from][]{Mahadevan96_AverageRewardReinforcementLearningFoundationsAlgorithmsAndEmpiricalResults}.}
\end{figure}

\subsection{Optimality Criteria}
\label{subsec:More_Refined_Optimality_Criteria}

The aforementioned and informally introduced notions of gain- and bias-optimality are formalised and
generalised to \(n\)-discount-optimality.

\begin{definition}
  Due to \citet{Veinott69} for MDPs a policy \(\polopt\) is \emph{\(n\)-discount-optimal} for
  \(n=-1,0,1,\ldots\) for all states \(s \in \States\) with discount factor
  \(\gamma\) % and \(V_{\gamma}^{\pol}(s)\) being the value function as defined above
  %
  if and only if
  \begin{align*}
    \lim_{\gamma \to 1}(1-\gamma)^{-n}\ (V_{\gamma}^{\polopt}(s) - V_{\gamma}^{\pol}(s)) \geqslant 0 \tpkt
  \end{align*}
\end{definition}

Only if a policy is \(n\)-discount-optimal for all \(n < m\) it can be \(m\)-discount
optimal \citep{Puterman94,Veinott69}.
%
If a policy is \(\infty\)-discount-optimal then it is said to be Blackwell-optimal
\citep{Blackwell62}.
% weakening the argument!!!
That is, for Blackwell-optimal policies \(\polopt\) there exists a discount factor \(\gammaopt < 1\)
such that \(V_{\gamma}^{\polopt}(s) \geqslant V_{\gamma}^{\pol}(s)\) for all
\(\gamma \geqslant \gammaopt\) and under all policies \(\pol\)
\citep{Mahadevan96_SensitiveDiscountOptimalityUnifyingDiscountedAndAverageRewardReinforcementLearning,Blackwell62}.
Informally that means there exists a discount factor \(<1\) which finds Blackwell-optimal policies.
However, (i) in more complex, i.e.\@ real world, MDPs this value can be arbitrary close to \(1\) and
(ii) the difference in state values may be very small, which (iii) due to the need of state value
function approximation likely causes errors when choosing among different actions in the discounted
framework.

For the case of \(n=-1\) this lead to gain-optimality
\(\avgrew^{\polopt}(s) - \avgrew^{\pol}(s) \geqslant 0\) for all policies \(\pol\) and states \(s\)
\citep{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies}. In
the case of \(n=0\) it describes bias-optimality with \(V^{\polopt}(s) - V^{\pol}(s) \geqslant 0\)
for all policies \(\pol\) and states \(s\)
\citep{Mahadevan96_AnAveragerewardReinforcementLearningAlgorithmForComputingBiasoptimalPolicies}.
%

@@

where the error term \(e_{\gamma}^{\pol}(s)\) exists of infinitely many subterms and
\citet{Puterman94} shows that \(\lim_{\gamma \to 1} e_{\gamma}^{\pol}(s) = 0\). Note how the first
term depending on the average reward \(\avgrew^{\pol}(s)\) converges to infinity as \(\gamma\)
increases.

However, an important insight is the connection between \(n\)-discount-optimality and the Laurent
series expansion of \(V_{\gamma}^{\pol}(s)\). Each addend corresponds to one step in
\(n\)-discount-optimality. That is for \((-1)\)-discount-optimality the average reward must be
maximised.
%
Then for \(0\)-discount-optimality the agent has to choose the highest bias values.
%
Finally \(n\)-discount-optimality for \(n \geqslant 1\) requires to maximise the error term
\(e_{\gamma}^{\pol}(s)\) which only exists as \(\gamma\) is strictly less than \(1\). Thus this term
incorporates the number of expected steps and their corresponding reward on the path to the
Blackwell-optimal policy. That is, it optimises the expected reward collected to reach the Blackwell
optimal policy according to the occurrence on the paths, where shorter paths and those which collect
the rewards sooner are preferred.
